Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-15 submission/manuscript.Rmd
shigeki abstract edits

It is currently unclear why mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs). While some chromatin regulators are important for neuron-specific gene expression, loss of repressive chromatin regulators can also lead to the aberrant transcription of tissue-specific genes outside of their intended context. Because very few studies have characterized the these aberrant, tissue-specific transcripts, the mechanisms behind their dysregulation and their functional consequences are largely unknown. Here, we explored this cellular identity crisis in NDDs by investigating the ectopic expression of germline (e.g. sperm and egg) genes in mice lacking lysine demethylase 5c (KDM5C, also known as JARID1C or SMCX), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3). We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key behavioral phenotypes of Claes-Jensen X-linked intellectual disability (MRXSCJ, OMIM: 300534), aberrantly expresses many liver, muscle, ovary, and testis genes within the amygdala and hippocampus. Many genes expressed in the _Kdm5c_-KO brain were typically unique to germ cells, indicating an erosion of the soma-germline boundary. Using germ cell-depleted RNA sequencing datasets, we curated a a list of germline-enriched genes for unbiased characterization of KDM5C in germline gene repression. Germline genes are typically decommissioned in somatic lineages in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_. Germline gene dysregualtion is sexually dimorphic, as female _Kdm5c_-KO EpiLCs were more prone to germline gene dysregulation than knockout males. We found KDM5C acts at a subset of germline gene promoters in EpiLCs to promote the initial placement of DNA methylation at CpG islands. However, germline genes can also become activated in _Kdm5c_-KO cells independent of direct KDM5C regulation. Intriguingly, many of the germline transcripts expressed in the _Kdm5c_-KO brain were late-stage spermatogenesis genes that are not directly bound by KDM5C. This suggests germline developmental programs can progress ectopically in the background of typical _Kdm5c_-KO development. Ultimately, this work characterizes a novel suppressor of germline gene transcription and links impaired soma-vs-germline demarcation to a chromatin-linked neurodevelopmental disorder.

>>>>

It is currently unclear why mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs). Loss of repressive chromatin regulators can lead to the aberrant transcription of tissue-specific genes outside of their intended context, however the mechanisms behind their dysregulation and their functional consequences are largely unknown. Here, we explored this cellular identity crisis in NDDs by investigating the role of lysine demethylase 5c (KDM5C, also known as JARID1C or SMCX), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3), in tissue identity. We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key behavioral phenotypes of Claes-Jensen X-linked intellectual disability, aberrantly expresses many liver, muscle, ovary, and testis genes within the amygdala and hippocampus. Gonad-enriched genes expressed in the _Kdm5c_-KO brain were typically unique to germ cells, indicating an erosion of the soma-germline boundary. We then curated a list of germline-enriched genes for unbiased characterization of KDM5C in germline gene repression. Germline genes are typically decommissioned in somatic lineages in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_. Germline gene dysregualtion was sexually dimorphic, as female _Kdm5c_-KO EpiLCs were more prone to germline gene misexpression than knockout males. We found KDM5C represses germline-specific transcription by binding to a subset of germline gene promoters in EpiLCs to facilitate DNA methylation of CpG islands. However, germline genes, particularly late-stage spermatogenesis genes, can also become activated in _Kdm5c_-KO cells independent of direct KDM5C binding. This suggests germline developmental programs can progress ectopically in the background of typical _Kdm5c_-KO development, due to downstream activation by germline transcription factors. These results define KDM5C's role in germline gene suppression and potentially link impaired soma-vs-germline demarcation to a chromatin-based neurodevelopmental disorder.

- Not sure if I like the last sentence, I think we should focus on what we did do versus what we could do in the future
